---
title: "Julia-R Integration"
author: "Zechariah Meunier"
date: "`r Sys.Date()`"
output: html_document
---

Load libraries
```{r}
library(JuliaCall)
library(diffeqr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

Set up Julia with local filepath
```{r}
julia_setup(JULIA_HOME = "C:/Users/zdmeu/AppData/Local/Programs/Julia-1.9.3/bin")

de <- diffeq_setup()
```

Test with Lotka-Volterra model
```{r}
#Define function for prey (du1) and predator (du2)
f <- function(u, p, t) {
    du1 <- p[1] * u[1] + p[2] * u[1] * u[2]
    du2 <- p[3] * u[2] + p[4] * u[1] * u[2]
    return( c(du1, du2) )
}

#Set parameters (p), initial populations (u0), and time span (tspan)
p <- c(0.1, -0.01, -0.05, 0.001)
u0 <- c(50, 15)
tspan <- c(0, 240)
```

Assign the time point at which a solution shall be returned (to make the plotted function smoother) and call Julia’s Differential Equation solver
```{r}
t <- seq(0, 240, by=2)

prob <- de$ODEProblem(f, u0, tspan, p)
sol <- de$solve(prob, reltol = 1e-08, saveat = t)

#Convert Julia object into an R data frame
y <- sapply(sol$u,identity)

#Wrangle y in tidy format
y2 <- data.frame(t(y)) %>%
  mutate(Time = t) %>%
  pivot_longer(cols = c(X1, X2),
               names_to = "Species",
               values_to = "Abundance")

y3 <- y2 %>%
  mutate(Abundance = Abundance - 100)
```

Plot the predator and prey populations with base plot
```{r}
par(mar=c(4, 2, 2, 1))
plot(0, 0, type = 'n', xlim = c(0, 250), ylim = c(0, 100),
     xlab = 'time', ylab = '',
     main = 'Solution to Lotka-Volterra equations')
grid()

lines(t, y[1, ], col = 3, lwd = 1.5)
lines(t, y[2, ], col = 2, lwd = 1.5)
```


Plot the predator and prey populations with ggplot2
```{r}
ggplot(data = y2, aes(x = Time, y = Abundance, group = Species)) +
  geom_line()
```

Plot the predator and prey populations with package function
```{r}
source("../src/plotting.R")
series_plot(data = y2, x = Time, y = Abundance, group = Species)
predictions_plot(data = y2, predictions = y3, x = Time, y = Abundance, group = Species)
```

