---
title: "emude at ESA 2025"
author: "Zechariah Meunier"
date: '`r Sys.Date()`'
output: html_document
---
# Setup

Load dependencies
```{r}
library(JuliaCall)
```

Set up Julia with local filepath
```{r}
julia_setup(JULIA_HOME = "C:\\Users\\zdmeu\\.julia\\juliaup\\julia-1.11.4+0.x64.w64.mingw32\\bin")
julia_library("UniversalDiffEq")
```

Load source code

```{r, include=FALSE}
setwd("..")

# Source Julia files
jl_files <- list.files("src", pattern = "\\.jl$", full.names = TRUE)
lapply(jl_files, JuliaCall::julia_source)

# Source R files, excluding emude.R
r_files <- list.files("src", pattern = "\\.R$", full.names = TRUE)
r_files <- r_files[!basename(r_files) %in% c("emude.R")]
lapply(r_files, source)
```

# Case Study 1: Canadian Lynx-Hare Predator-Prey Cycles

## Data Processing

Load data

Dataset 1 is from https://gist.github.com/michaelosthege/27315631c1aedbe55f5affbccabef1ca
Dataset 2 is from https://jckantor.github.io/CBE30338/02.05-Hare-and-Lynx-Population-Dynamics.html

```{r}
lynxhare1 <- read.csv("Hare and Lynx Time Series 1.csv")
lynxhare1$Hare <- lynxhare1$Hare + 0.1
lynxhare1$Lynx <- lynxhare1$Lynx + 0.1
lynxhare2 <- read.csv("Hare and Lynx Time Series 2.csv")
```

Join data frames and see how they differ

```{r}
lynxhare <- right_join(lynxhare1, lynxhare2, by = "Year")

ggplot(data = lynxhare) +
  geom_abline(intercept = 0, slope = 1/1000) +
  geom_point(aes(x = Hare.x, y = Hare.y, color = "1")) +
  geom_point(aes(x = Lynx.x, y = Lynx.y, color = "2")) +
  scale_x_continuous(name = "Dataset 1 population (pelts)") +
  scale_y_continuous(name = "Dataset 2 population (thousand pelts)") +
  scale_color_manual(name = "",
                     labels = c("Hare","Lynx"),
                     values = c("dodgerblue","firebrick")) +
  theme_bw() +
  theme(legend.position = "top")
```


Plot time series
```{r}
ggplot(data = lynxhare1, aes(x = Year)) +
  geom_line(aes(y = Hare, color = "1")) +
  geom_point(aes(y = Hare, color = "1")) +
  geom_line(aes(y = Lynx, color = "2")) +
  geom_point(aes(y = Lynx, color = "2")) +
  scale_x_continuous(breaks = seq(1845, 1935, 5)) +
  scale_y_continuous(name = "Population (pelts)") +
  scale_color_manual(name = "",
                     labels = c("Hare","Lynx"),
                     values = c("dodgerblue","firebrick")) +
  theme_bw() +
  theme(legend.position = "top")

ggplot(data = lynxhare2, aes(x = Year)) +
  geom_line(aes(y = Hare, color = "1")) +
  geom_point(aes(y = Hare, color = "1")) +
  geom_line(aes(y = Lynx, color = "2")) +
  geom_point(aes(y = Lynx, color = "2")) +
    scale_x_continuous(breaks = seq(1845, 1935, 5)) +
  scale_y_continuous(name = "Population (thousand pelts)") +
  scale_color_manual(name = "",
                     labels = c("Hare","Lynx"),
                     values = c("dodgerblue","firebrick")) +
  theme_bw() +
  theme(legend.position = "top")
```

## Neural Ordinary Differential Equation (NODE)

Create a NODE model on data set 1 without rescaling the data. Then train this model and make predictions based on the original data frame.

```{r}
nodemodel1 <- NODE(data = lynxhare1, time_column_name = "Year")
train(nodemodel1)
predict(nodemodel1, test_data = lynxhare1)

julia_eval(paste0(nodemodel1, ".data_frame"))

```

